---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First part of assignment 1 - Getting and Cleaning the data:
```{r}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, "Course5Assignment1.zip")
unzip("Course5Assignment1.zip")
df <- read.csv("activity.csv", sep=",", na.strings = "NA", stringsAsFactors = F, colClasses = c(steps = "numeric", date = "character", interval = "numeric"))
df$date <- as.Date(df$date, "%Y-%m-%d")
```


## What is mean total number of steps taken per day?
Second part of assignment 1 - Reporting on the total number of steps:

- calculating the sum of steps per day
- plotting them into a histogram
- reporting the mean and median of the number of steps
```{r}
pivot_df <- aggregate(steps ~ date, df, sum)
head(pivot_df, 3)
barplot(pivot_df$steps, main = "Total Steps per Day", col="red", ylab="# of steps")
mean_df <- mean(pivot_df$steps)
print(paste("The MEAN of the total number of steps taken per day is ", mean_df,"(see green dotted line)."))
abline(h = mean_df, lwd = 3, lty = 3, col = "green")
median_df <- median(pivot_df$steps)
print(paste("The MEDIAN of the total number of steps taken per day is ", median_df,"(see blue dashed line)."))
barplot(pivot_df$steps, main = "Total Steps per Day", col="red", ylab="# of steps")
abline(h = median_df, lwd = 2, lty = 2, col = "blue")
```


## What is the average daily activity pattern?
Third part of assignment 1 - time series of average daily activities:

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
pivot_2_df <- aggregate(steps ~ interval, df, mean)
with(pivot_2_df, plot(interval, steps, type = "l"))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_interval <- pivot_2_df[which.max(pivot_2_df$steps),]
print(paste("The interval with the maximum number of average steps is", max_interval$interval,"and has", round(max_interval$steps, 1),"steps."))
```


## Imputing missing values

```{r}
```


## Are there differences in activity patterns between weekdays and weekends?

```{r}
```
